---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Include libraries
```{r}
library (lubridate)
library (ggplot2)
library (stringr)
library (dplyr)
library (knitr)
library (impute)

```

## Download data
```{r download}
zipurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destzip <- "accelerometerdata.zip"
destfile <- "."
download.file(url=zipurl, destfile=destzip)
unzip(destzip)
```

## Load data
```{r load}
df <- read.csv("activity.csv")
```

### Perform date-time processing; not used eventually
```{r datetime}
#df$interval <- stringr::str_pad(as.character(df$interval),4,pad='0')
#df$interval <- paste(substr(df$interval,1,2), ':', substr(df$interval,3,4), sep="")
#df$date <- strptime(paste(df$date, df$interval), "%Y-%m-%d %H:%M")
```


## What is mean total number of steps taken per day?
```{r}
stepsDay <- tapply(df$steps, df$date, sum, na.rm=TRUE)
qplot(stepsDay, xlab='Total Steps per Day', ylab='Frequency', binwidth = 500)
stepsMean <- mean(stepsDay)
stepsMedian <- median(stepsDay)
```

* Mean number of steps: `r stepsMean`
* Median number of steps: `r stepsMedian`

## What is the average daily activity pattern?
```{r}
avgStepsEachBlock <- aggregate(x=list(meanSteps=df$steps), by=list(interval=df$interval), FUN=mean, na.rm=TRUE)
```
## Time series plot of number of steps taken at different time intervals
```{r}
ggplot(data=avgStepsEachBlock, aes(x=interval, y=meanSteps)) + geom_line() + xlab("5 minute interval") +
    ylab("Average number of steps") 

```

## Interval having the max number of steps on average
```{r}
maxSteps <- which.max(avgStepsEachBlock$meanSteps)
timeMaxSteps <- avgStepsEachBlock[maxSteps, 'interval']
timeMaxSteps <- stringr::str_pad(as.character(timeMaxSteps),4,pad='0')
timeMaxSteps <- paste(substr(timeMaxSteps,1,2),':', substr(timeMaxSteps,3,4), sep="")


```
* Max Steps at: `r timeMaxSteps`

## Imputing missing values
```{r}
numNA <- length(which(is.na(df$steps)))  #number of missing values
```
Number of missing values: `r numNA`

## Devise a strategy for filling in all of the missing values in the dataset
```{r}
#impute function just didn't work; tried all options
#finally dropped the idea and calculated median for intervals and used that instead
dfImputed <- df
narows <- is.na(dfImputed$steps)
avgInterval <- tapply(dfImputed$steps, dfImputed$interval, median, na.rm=TRUE, simplify = TRUE)
dfImputed$steps[narows] <- avgInterval[as.character(dfImputed$interval[narows])]

#dfImputed[is.na(dfImputed$steps),]$steps <- stepsMean/24/12  #daily mean per 5 mins; too simplistic
```



## What is mean and median total number of steps taken per day (after impute)?
```{r}
stepsImpDay <- tapply(dfImputed$steps, dfImputed$date, sum, na.rm=TRUE)
qplot(stepsImpDay, xlab='Total Steps per Day', ylab='Frequency', binwidth = 500)
stepsImpMean <- mean(stepsImpDay)
stepsImpMedian <- median(stepsImpDay)
```

* Mean number of steps (after impute): `r stepsImpMean`
* Median number of steps (after impute): `r stepsImpMedian`


## Are there differences in activity patterns between weekdays and weekends?
```{r}
dfImputed$dateType <-  ifelse(as.POSIXlt(dfImputed$date)$wday %in% c(0,6), 'weekend', 'weekday')

avgdfImputed <- aggregate(steps ~ interval + dateType, data=dfImputed, mean)
ggplot(avgdfImputed, aes(x=interval, y=steps)) + geom_line() + facet_grid(dateType ~ .) + xlab("5 minute interval") + ylab("avg number of steps")
```
